A Formal Model of Checked C

### 2023-01-14 sandp23/formal.tex

\subsection{Type Checking}
\label{sec:typechecking}
Our type checker restricts the usage of tainted and checked pointer types to ensure that tainted pointers do not affect checked types, along with enforcing~\checkedc typing rules~\cite{li22checkedc}.
A few selected type rules are presented in~\fig{fig:type-system-1}.
% 
\mz{Move the figure here.}
% 
The complete set of typing rules and special handling of (NT)-arrays are provided in~\apdx{sec:literal-pointer-typing}.

Each typing judgment has the form $\Gamma;\Theta\vdash_m e : \tau$,
which states that in a type environment $\Gamma$ (mapping variables to
their types) and a predicate environment $\Theta$ (mapping integer-typed
variables to Boolean predicates), expression $e$ will have type $\tau$ if evaluated
in context mode $m$.

In \lang, the type equability ($=_{\Theta}$) and subtype ($\sqsubseteq$) relations are given in \Cref{fig:checkc-subtype}.
We provide some example descriptions here.
Type equality $\tau=_{\Theta}\tau'$
is a type construct equivalent relation defined by the bound equality ($=_{\Theta}$) in (NT-)array pointer types
and the alpha equivalence of two function types;
i.e., two (NT-)array pointer types $\tallarrayb{\bvar}{\tau} $ and $ \tallarrayb{\bvar'}{\tau'}$ are equivalent, if 
$\bvar =_{\Theta} \bvar'$ and $\tau=_{\Theta}\tau'$; two function types 
$\tfun{\overline{x}}{\overline{\tau}}{\tau} $ and $ \tfun{\overline{y}}{\overline{\tau'}}{\tau'}$
are equivalent, if we can find a same length (as $\overline{x}$ and $\overline{y}$) variable list $\overline{z}$ that is substituted for $\overline{x}$ and $\overline{y}$ in $\overline{\tau} \to {\tau}$ and $\overline{\tau'} \to {\tau'}$, resp.,
and the substitution results are equal.
\aravind{For the above paragraph, give concrete example, For example: array is a subtype of ptr, etc}

The type checker rules are designed such that any well-typed (\ie follows all the rules) is isolated by sandboxed code. We present the corresponding formal proof in~\apdx{}.

One of the important type cheker rules is disallowing unsafe interactions between tainted and untainted (c/u) pointers.
\aravind{Explain about the LET rule}
%
Rules \textsc{T-Let} and \textsc{T-LetInt} in in \Cref{fig:type-system-1} type a $\elettext$ expression, which also admits
type dependency. 
In particular, the result of evaluating a $\elettext$ expression
may have a type that refers to one of its bound variables (e.g., if
the result is a checked pointer with a variable-defined bound). 
If so, we must substitute away this variable once it goes out of scope (\textsc{T-LetInt}). 
Note that we restrict the expression $e_1$ to syntactically match the
structure of a Bounds expression $b$ (see Fig.~\ref{fig:checkc-syn}).
Rule \textsc{T-RetInt} types a $\erettext$ expression when $x$ is of type $\tint$.
$\erettext$ does not appear in source programs but is introduced by the semantics when
evaluating a let binding (rule \textsc{S-Let} in
Fig.~\ref{fig:semantics}). 
%\liyi{why? }
% After the evaluation of a let binding a variable $x$ concludes,
%we need to restore any prior binding of $x$, which is either
%$\bot$ (meaning that there is no $x$ originally) or some value
%$\evalue{n}{\tau}$.

\aravind{We should explain how our type checker restricts unsafe interactions between checked and tainted pointer types.}

\begin{minted}[xleftmargin=30pt, mathescape, escapeinside=||, fontsize=\footnotesize]{c}
a -> checked pointer
b -> Tainted pointer
c -> Unchecked pointer.

_Ptr<int> a;
_t_Ptr<int> b;
int *c;

a = b; // Not allowed.
b = a; // Not allowed.
\end{minted}

The \textsc{T-CastPtr} rule in \Cref{fig:type-system-1}
permits casting from an expression of type $\tau'$ to a checked pointer when
$\tau' \sqsubseteq \tptr{\tau}{\cmode}$. This subtyping relation
$\sqsubseteq$ is built on the type equality ($\tau =_{\Theta} \tau'\Rightarrow\tau \sqsubseteq_{\Theta} \tau'$). 
The rule  ($0\le b_l \wedge b_h \le 1 \Rightarrow \tptr{\tau}{m}\sqsubseteq
\tarrayptr{b_l}{b_h}{\tau}{m}$) permits treating a singleton
pointer as an array pointer with $b_h\le 1$ and $0 \le b_l$.
Two function pointer types are subtyped ($\tptr{\tfun{\overline{x}}{\overline{\tau}}{\tau}}{\xi} \sqsubseteq_{\Theta} \tptr{\tfun{\overline{x}}{\overline{\tau'}}{\tau'}}{\xi}$), 
if the output type are subtyped ($\tau\sqsubseteq_{\Theta}\tau'$) and the argument types are reversely subtyped ($\overline{\tau'}\sqsubseteq_{\Theta}\overline{\tau}$).

\aravind{Explain how the cast rule allows/prevents the casts.}

\begin{minted}[xleftmargin=30pt, mathescape, escapeinside=||, fontsize=\footnotesize]{c}
a -> checked pointer
b -> Tainted pointer
c -> Unchecked pointer.

_Ptr<int> a;
_t_Ptr<int> b;
int *c;

a = b; // Not allowed.
b = a; // Not allowed.

a = (_Ptr<int>)c; // This is okay.
a = (_Ptr<int>)b; // Not allowed.
\end{minted}

>>>>

\subsection{Typing and Semantics}
\label{sec:typechecking}

The \lang system is a flow-sensitive, gradual type one that generates additional dynamic checks that are inserted in the typing checking stage and executed in the semantic evaluation stage.
Our type checker restricts the usage of tainted and checked pointer types to ensure that tainted pointers do not affect checked types, along with enforcing~\checkedc typing rules~\cite{li22checkedc}.

As partly shown in \Cref{fig:type-system-1} (labeled as \rulelab{T-}$X$),
Each typing judgment has the form $\Gamma;\Theta\vdash_m e : \tau$,
which states that in a type environment $\Gamma$ (mapping variables to
their types) and a predicate environment $\Theta$ (mapping integer-typed
variables to Boolean predicates), expression $e$ will have type $\tau$ if evaluated
in context mode $m$.
The operational semantics for \lang is defined as a small-step
transition relation with the judgment $ (\varphi,\heap,e)
\longrightarrow_m (\varphi',\heap',r)$, as partly shown in \Cref{fig:type-system-1} (labeled as \rulelab{S-}$X$).
 Here, $\varphi$ is a
\emph{stack} mapping from variables to values $\evalue{n}{\tau}$ and
$\heap$ is a \emph{heap} that is partitioned into two parts ($\cmode$ and $\umode$ heaps), each of which
maps addresses (integer literals) to values $\evalue{n}{\tau}$.
The complete set of typing rules and special handling of (NT)-arrays are provided in~\apdx{sec:literal-pointer-typing}.

\myparagraph{Modes, Static Casting, and Subtyping}
In \lang, Context modes $m$ appearing in a type rule determine the code region 
that permits pointer dereferences and value-assignments, depending on the pointer modes.
We define a three point lattice $\xi_1 \le \xi_2$ to describe such permission, where $\tmode \le \xi$ and $m \le m$.
This means that a $\tmode$ pointer can be dereferenced and value-assigned in any region, while $\cmode$ and $\umode$ pointers can perform such operations in $\cmode$ and $\umode$ regions, respectively.
Pointer modes are also useful in determining if a nested pointer has a valid type.
For example, in a nested pointer $\tptr{(... \tptr{\tau}{\xi_2} ...)}{\xi_1}$, we require $\xi_2\le \xi_1$ to maintain non-exposure.

\lang also has static casting operations. As described in rule \textsc{T-CastPtr},
an pointer typed expression of type $\tptr{\tau_1}{\xi_1}$ can be casted to another pointer type \tptr{\tau_2}{\xi_2},
if \tptr{\tau_1}{\xi_1} subtypes ($\sqsubseteq_{\Theta}$) to \tptr{\tau_2}{\xi_2} $\tptr{\tau_1}{\xi_1} \sqsubseteq \tptr{\tau_2}{\xi_2}$.
In \lang, except that we can cast a $\tmode$ mode pointer to a $\umode$ mode one, all subtyping relations are between two types with the same mode, meaning that $\xi_1$ and $\xi_2$ above are mostly the same and the above mode lattice ($\le$) has no business with subtyping.

\begin{minted}[xleftmargin=30pt, mathescape, escapeinside=||, fontsize=\footnotesize]{c}
//_Ptr<int> a; _t_Ptr<int> b; int *c;
c = (int *)b; // This is okay.
a = (_Ptr<int>)b; // Not allowed.
\end{minted}

In the above example, a $\tmode$ mode pointer can be cast to $\umode$ mode but casting $\tmode$ mode to $\cmode$ mode is disallowed.
The complete subtyping relation was described in \Cref{fig:checkc-subtype}.
\ignore{
For example, the rule ($0\le b_l \wedge b_h \le 1 \Rightarrow \tptr{\tau}{\xi}\sqsubseteq
\tarrayptr{b_l}{b_h}{\tau}{\xi}$) permits treating a singleton
pointer as an array pointer with $b_h\le 1$ and $0 \le b_l$.
Two function pointer types are subtyped ($\tptr{\tfun{\overline{x}}{\overline{\tau}}{\tau}}{\xi} \sqsubseteq_{\Theta} \tptr{\tfun{\overline{x}}{\overline{\tau'}}{\tau'}}{\xi}$), 
if the output type are subtyped ($\tau\sqsubseteq_{\Theta}\tau'$) and the argument types are reversely subtyped ($\overline{\tau'}\sqsubseteq_{\Theta}\overline{\tau}$).}
Notice that \texttt{let} statements are immutable in \lang, so the following code is not possible, because variables \code{x} and \code{y} must have the same type in \lang.

\begin{minted}[xleftmargin=30pt, mathescape, escapeinside=||, fontsize=\footnotesize]{c}
//_Ptr<int> x;  _t_Ptr<int> y;
x = y; // Not allowed.
\end{minted}
